Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die Einhaltung der Echtzeit

### 2024-06-03 Masterarbeit_Robotik_Pamuk.tex
secnumdepth, tocdepth
minor
latency from source

%
% FH Technikum Wien
% !TEX encoding = UTF-8 Unicode
%
% Erstellung von Master- und Bachelorarbeiten an der FH Technikum Wien mit Hilfe von LaTeX und der Klasse TWBOOK
%
% Um ein eigenes Dokument zu erstellen, müssen Sie folgendes ergänzen:
% 1) Mit \documentclass[..] einstellen: Master- oder Bachelorarbeit, Studiengang und Sprache
% 2) Mit \newcommand{\FHTWCitationType}.. Zitierstandard festlegen (wird in der Regel vom Studiengang vorgegeben - bitte erfragen)
% 3) Deckblatt, Kurzfassung, etc. ausfüllen
% 4) und die Arbeit schreiben (die verwendeten Literaturquellen in Literature.bib eintragen)
%
% Getestet mit TeXstudio mit Zeichenkodierung ISO-8859-1 (=ansinew/latin1) und MikTex unter Windows
% Zu beachten ist, dass die Kodierung der Datei mit der Kodierung des paketes inputenc zusammen passt!
% Die Kodierung der Datei twbook.cls MUSS ANSI betragen!
% Bei der Verwendung von UTF8 muss dnicht nur die Kodierung des Dokuments auf UTF8 gestellt sein, sondern auch die des BibTex-Files!
%
% Bugreports und Feedback bitte per E-Mail an latex@technikum-wien.at
%
% Versionen
% *) V0.7: 9.1.2015, RO: Modeline angepasst und verschoben
% *) V0.6: 10.10.2014, RO: Weitere Anpassung an die UK
% *) V0.5: 8.8.2014, WK: Literaturquellen überarbeitet und angepasst
% *) V0.4: 4.8.2014, WK: Initalversion in SVN eingespielt
%
\documentclass[MMR,Master,english]{twbook}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%
% Hier biblatex & Biber konfigurieren; Vergessen Sie nicht, dass Sie biber verwenden müssen um eine Bibliothek zu erzeugen
%
\usepackage[backend=biber, style=numeric, sorting=none]{biblatex}
\addbibresource{Literature.bib}
%
% Bei Bedarf bitte hier die Syntax-Highlightings anpassen
%
% own 
\usepackage{hyperref}
\usepackage{float}
\usepackage{comment}
\usepackage{csquotes}
\usepackage{pgfplots}
\usepackage{soul}
\usepackage{color}
\usepackage{xcolor}  % Required for custom colors
\sethlcolor{blue}
%
% Platzierungsoptionen 
%
\usepackage[final]{listings}
\lstset{captionpos=b, numberbychapter=false,caption=\lstname,frame=single, numbers=left, stepnumber=1, numbersep=2pt, xleftmargin=15pt, framexleftmargin=15pt, numberstyle=\tiny, tabsize=3, columns=fixed, basicstyle={\fontfamily{pcr}\selectfont\footnotesize}, keywordstyle=\bfseries, commentstyle={\color[gray]{0.33}\itshape}, stringstyle=\color[gray]{0.25}, breaklines, breakatwhitespace, breakautoindent}
\lstloadlanguages{[ANSI]C, C++, [gnu]make, gnuplot, Matlab}

@@

\clearpage
\chapter{Salamander 4}\label{cha:salamander4}
This chapter briefly describes the Salamander 4 operating system by SIGMATEK.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/max_latency_hardware.png}
	\caption[Latency hardware]{Latency hardware}
	\label{fig:max_latency_hardware}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/0hardware/gnuplot_max_latency_hardware.png}
	\caption[gnuplot latency hardware]{gnuplot latency hardware}
	\label{fig:gnuplot_max_latency_hardware}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/max_latency_default.png}
	\caption[Latency no taskset]{Latency no taskset}
	\label{fig:max_latency_default}
\end{figure}

Figure \ref{fig:gnuplot_max_latency_default} depicts the variation in latency over the course of said time.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/1default/gnuplot_max_latency_default.png}
	\caption[gnuplot latency hardware]{gnuplot latency hardware}
	\label{fig:gnuplot_max_latency_default}
\end{figure}

@@

\clearpage
\chapter{Real-Time Performance Tuning}\label{cha:real-time_tuning}
\section{BIOS Configurations}
\begin{table}[h]
	\centering
	\caption{BIOS Configurations}
	\begin{tabular}{|l|l|}
	\hline
	\textbf{Option} & \textbf{Status} \\
	\hline
	Hyper Threading & Disabled \\
	\hline
	Intel SpeedStep® & Disabled \\
	\hline
	Intel® Speed Shift Technology & Disabled \\
	\hline
	C States & Disabled \\
	\hline
	VT-d & Enabled \\
	\hline
	\end{tabular}
	\end{table}

\section{Kernel Configurations}

@@

\begin{itemize}
	\item \textbf{isolcpus=4}: This isolates CPU 4 from the general scheduler, meaning no process will be scheduled to run on this CPU unless it is explicitly assigned.
	\item \textbf{rcu\_nocbs=4}: Moves the RCU (Read-Copy-Update) callback handling off of CPU 4.
	\item \textbf{nohz\_full=4}: Makes CPU 4 tickless, reducing the timer interrupts and thus can improve performance for specific applications.
	\item \textbf{default\_hugepagesz=1G, hugepagesz=1G, hugepages=8}: Sets the default huge page size to 1GB, and reserves 8 huge pages at boot.
	\item \textbf{intel\_iommu=on}: Enables Intel's IOMMU (Input/Output Memory Management Unit) for hardware that supports it.
	\item \textbf{rdt=l3cat}: Enables the L3 Cache Allocation Technology.
	\item \textbf{nmi\_watchdog=0}: Disables the NMI watchdog, which can free up a bit of system resources.
	\item \textbf{idle=poll}: Changes the idle loop of the CPU to actively poll.
	\item \textbf{clocksource=tsc, tsc=reliable}: Sets the clocksource to TSC (Time Stamp Counter) and marks it as reliable.
	\item \textbf{noht}: Disables Hyper-Threading.
	\item \textbf{audit=0}: Disables the Linux audit system.
	\item \textbf{skew\_tick=1}: Enables a mode to reduce timer interrupt overhead.
	\item \textbf{intel\_pstate=disable}: Disables the intel\_pstate driver.
	\item \textbf{intel.max\_cstate=0, intel\_idle.max\_cstate=0, processor.max\_cstate=0, processor\_idle.max\_cstate=0}: Disables deeper C-states, which can help reduce latency.
	\item \textbf{nosoftlockup}: Disables the soft lockup detector.
	\item \textbf{nohz=on}: Enables tickless operation system-wide.
	\item \textbf{no\_timer\_check}: Disables the check for broken timer interrupt sources.
	\item \textbf{nospectre\_v2 spectre\_v2\_user=off}: Disables mitigations for the Spectre v2 vulnerability.
	\item \textbf{kvm.kvmclock\_periodic\_sync=N, kvm\_intel.ple\_gap=0}: These are KVM (Kernel-based Virtual Machine) related parameters.
	\item \textbf{irqaffinity=0}: Sets the default IRQ affinity.
\end{itemize}

\section{Host OS Configurations}
\subsection{CPU isolation}

@@

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
	\begin{lstlisting}[name={User and Kernel Tasks},label={output:pse},escapeinside={(*@}{@*)}]
		sigma_ibo@sigma-ibo:~$ cat /sys/devices/system/cpu/isolated
		4
		sigma_ibo@sigma-ibo:~$ ps axHo psr,pid,lwp,args,policy,nice,rtprio | awk '$1 == 4'
		4      38      38 [cpuhp/4]                   TS    0      -
		4      39      39 [idle_inject/4]             FF    -     50
		4      40      40 [migration/4]               FF    -     99
		4      41      41 [ksoftirqd/4]               TS    0      -
		4      42      42 [kworker/4:0-events]        TS    0      -
		4      43      43 [kworker/4:0H-events_highpr TS  -20      -
		4     505     505 [irq/189-iwlwifi:queue_1]   FF    -     50
		4     519     519 [irq/203-iwlwifi:exception] FF    -     50
		4    3008    3008 [kworker/4:1H]              TS  -20      -
		4    3022    3022 [kworker/4:2-rcu_gp]        TS    0      -	  
\end{lstlisting}
\end{minipage}

\subsection{Interrupt Requests Handling}\label{sec:irq_handling}
Once the CPUs were isolated, Interrupt Requests Handling was the next step. Interrupt Requests are used to send a signal to the CPU, prompting it to 'interrupt' its current task and divert its attention to another task. This allows hardware devices to communicate with the CPU through frequent context switches, which can lead to performance degradation, especially in high-performance computing or real-time scenarios. To mitigate this, the IRQs needed to be removed from the isolated CPUs. This was done by manipulating a file in the proc filesystem, namely \texttt{/proc/irq/<IRQ>/smp\_affinity}. The value in the \texttt{smp\_affinity} file is a bitmask in hexadecimal format. Each bit in this mask corresponds to a CPU in the system. The least significant bit (LSB) on the right corresponds to the first CPU (CPU0), and the significance increases towards the left until CPU19. In a system with 20 CPUs, if every CPU was reserved for one IRQ, the value for \texttt{smp\_affinity} would be FFFFF. The script in code \ref{script:smp_affinity} was written to check and log the distribution of Interrupt Requests across each CPU in Salamander 4.

@@

\noindent By changing the values of the \texttt{smp\_affinity} files of the respective IRQs, the assignment of IRQs was controlled so that they would not be handled by the isolated CPU. This reduced the interruptions caused by IRQs and the isolated CPUs were able to focus more on their assigned tasks.

\subsection{Disable simultaneous multithreading}
Disable
\subsection{Disable dynamic frequency scaling}
\subsection{Disable RT throttling}
\subsection{No unexpected RT processes are running on your system}
\subsection{IRQ affinity}
\subsection{RCU CPU offloading}
\subsection{Suppress rcu cpu stall}
\subsection{Maybe?}
\subsection{Start QEMU normally and give all QEMU threads rt-priority}
\subsection{Kill all running user processes}
\subsection{Set CPU Affinity for systemd services}
\subsection{Cache Isolation for CPU and GPU}
\subsection{Set CPU Affinity of IRQ thread to CPU 0}
\subsection{Set Device Driver Work Queue to CPU 0}
\subsection{Disable Machine Check}
\subsection{Stop Certain Services}

\section{QEMU-KVM Configurations}
\subsection{Tune lapic timer advance}
\subsection{Set QEMU options for real-time VM}
\subsection{Set CPU affinity and scheduling policy of QEMU CPU threads}
\subsection{Passthrough PCI devices into the VM}

\section{Guest OS Configurations}

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualisation},label={script:qemu_configuration_optimizations}]
    #!/bin/sh

@@

\begin{comment}
\chapter{To Include}\label{cha:to_include}
\section{Latency Comparison}
In the initial phase, a comparative latency analysis was conducted between the hardware version and the virtualized version of Salamander 4. For this purpose, the latency tool of the Xenomai test suite was used. The latency was measured under two conditions, idle and CPU-stressed. The goal was to optimize the latency of the virtualisation of Salamander 4 OS to closely match that of the bare metal version.

>>>>

%
% FH Technikum Wien
% !TEX encoding = UTF-8 Unicode
%
% Erstellung von Master- und Bachelorarbeiten an der FH Technikum Wien mit Hilfe von LaTeX und der Klasse TWBOOK
%
% Um ein eigenes Dokument zu erstellen, müssen Sie folgendes ergänzen:
% 1) Mit \documentclass[..] einstellen: Master- oder Bachelorarbeit, Studiengang und Sprache
% 2) Mit \newcommand{\FHTWCitationType}.. Zitierstandard festlegen (wird in der Regel vom Studiengang vorgegeben - bitte erfragen)
% 3) Deckblatt, Kurzfassung, etc. ausfüllen
% 4) und die Arbeit schreiben (die verwendeten Literaturquellen in Literature.bib eintragen)
%
% Getestet mit TeXstudio mit Zeichenkodierung ISO-8859-1 (=ansinew/latin1) und MikTex unter Windows
% Zu beachten ist, dass die Kodierung der Datei mit der Kodierung des paketes inputenc zusammen passt!
% Die Kodierung der Datei twbook.cls MUSS ANSI betragen!
% Bei der Verwendung von UTF8 muss dnicht nur die Kodierung des Dokuments auf UTF8 gestellt sein, sondern auch die des BibTex-Files!
%
% Bugreports und Feedback bitte per E-Mail an latex@technikum-wien.at
%
% Versionen
% *) V0.7: 9.1.2015, RO: Modeline angepasst und verschoben
% *) V0.6: 10.10.2014, RO: Weitere Anpassung an die UK
% *) V0.5: 8.8.2014, WK: Literaturquellen überarbeitet und angepasst
% *) V0.4: 4.8.2014, WK: Initalversion in SVN eingespielt
%
\documentclass[MMR,Master,english]{twbook}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%
% Hier biblatex & Biber konfigurieren; Vergessen Sie nicht, dass Sie biber verwenden müssen um eine Bibliothek zu erzeugen
%
\usepackage[backend=biber, style=numeric, sorting=none]{biblatex}
\addbibresource{Literature.bib}
%
% Bei Bedarf bitte hier die Syntax-Highlightings anpassen
%
% own 
\usepackage{hyperref}
\usepackage{float}
\usepackage{comment}
\usepackage{csquotes}
\usepackage{pgfplots}
\usepackage{soul}
\usepackage{color}
\usepackage{xcolor}  % Required for custom colors
\sethlcolor{blue}
\setcounter{secnumdepth}{3} % in text 
\setcounter{tocdepth}{3} % contents

%
% Platzierungsoptionen 
%
\usepackage[final]{listings}
\lstset{captionpos=b, numberbychapter=false,caption=\lstname,frame=single, numbers=left, stepnumber=1, numbersep=2pt, xleftmargin=15pt, framexleftmargin=15pt, numberstyle=\tiny, tabsize=3, columns=fixed, basicstyle={\fontfamily{pcr}\selectfont\footnotesize}, keywordstyle=\bfseries, commentstyle={\color[gray]{0.33}\itshape}, stringstyle=\color[gray]{0.25}, breaklines, breakatwhitespace, breakautoindent}
\lstloadlanguages{[ANSI]C, C++, [gnu]make, gnuplot, Matlab}

@@

\chapter{Salamander 4}\label{cha:salamander4}
This chapter briefly describes the Salamander 4 operating system by SIGMATEK.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/max_latency_hardware.png}
	\caption[Latency in hardware]{Latency in hardware}
	\label{fig:max_latency_hardware}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/0hardware/gnuplot_max_latency_hardware.png}
	\caption[Variation in latency of hardware]{Variation in latency of hardware}
	\label{fig:gnuplot_max_latency_hardware}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/max_latency_default.png}
	\caption[Latency with default settings]{Latency with default settings}
	\label{fig:max_latency_default}
\end{figure}

Figure \ref{fig:gnuplot_max_latency_default} depicts the variation in latency with default settings over the course of said time.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/1default/gnuplot_max_latency_default.png}
	\caption[Variation in latency with default settings]{Variation in latency with default settings}
	\label{fig:gnuplot_max_latency_default}
\end{figure}

@@

\chapter{Real-Time Performance Tuning}\label{cha:real-time_tuning}
\section{BIOS Configurations}
BIOS is firmware used to perform hardware initialization during the booting process. Some BIOS settings can have a significant impact on the real-time performance of the system [2]. The configurations in Table 4 are tuned for real-time performance.

\begin{table}[h]
	\centering
	\caption{BIOS Configurations}
	\begin{tabular}{|l|l|}
	\hline
	\textbf{Option} & \textbf{Status} \\
	\hline
	Hyper Threading & Disabled \\
	\hline
	Intel SpeedStep® & Disabled \\
	\hline
	Intel® Speed Shift Technology & Disabled \\
	\hline
	C States & Disabled \\
	\hline
	VT-d & Enabled \\
	\hline
	\end{tabular}
	\end{table}	
\section{Kernel Configurations}

@@

\begin{itemize}
	\item \textbf{isolcpus=4}: This isolates CPU 4 from the general scheduler, meaning no process will be scheduled to run on this CPU unless it is explicitly assigned.
	\item \textbf{rcu\_nocbs=4}: Moves the RCU (Read-Copy-Update) callback handling off of CPU 4.
	\item \textbf{nohz\_full=4}: Makes CPU 4 tickless, reducing the timer interrupts and thus can improve performance for specific applications.
	\item \textbf{default\_hugepagesz=1G, hugepagesz=1G, hugepages=8}: Sets the default huge page size to 1GB, and reserves 8 huge pages at boot.
	\item \textbf{intel\_iommu=on}: Enables Intel's IOMMU (Input/Output Memory Management Unit) for hardware that supports it.
	\item \textbf{rdt=l3cat}: Enables the L3 Cache Allocation Technology.
	\item \textbf{nmi\_watchdog=0}: Disables the NMI watchdog, which can free up a bit of system resources.
	\item \textbf{idle=poll}: Changes the idle loop of the CPU to actively poll.
	\item \textbf{clocksource=tsc, tsc=reliable}: Sets the clocksource to TSC (Time Stamp Counter) and marks it as reliable.
	\item \textbf{noht}: Disables Hyper-Threading.
	\item \textbf{audit=0}: Disables the Linux audit system.
	\item \textbf{skew\_tick=1}: Enables a mode to reduce timer interrupt overhead.
	\item \textbf{intel\_pstate=disable}: Disables the intel\_pstate driver.
	\item \textbf{intel.max\_cstate=0, intel\_idle.max\_cstate=0, processor.max\_cstate=0, processor\_idle.max\_cstate=0}: Disables deeper C-states, which can help reduce latency.
	\item \textbf{nosoftlockup}: Disables the soft lockup detector.
	\item \textbf{nohz=on}: Enables tickless operation system-wide.
	\item \textbf{no\_timer\_check}: Disables the check for broken timer interrupt sources.
	\item \textbf{nospectre\_v2, spectre\_v2\_user=off}: Disables mitigations for the Spectre v2 vulnerability.
	\item \textbf{kvm.kvmclock\_periodic\_sync=N, kvm\_intel.ple\_gap=0}: These are KVM (Kernel-based Virtual Machine) related parameters.
	\item \textbf{irqaffinity=0}: Sets the default IRQ affinity.
\end{itemize}
\section{Host OS Configurations}
\subsection{CPU isolation}

@@

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
	\begin{lstlisting}[name={User and Kernel Tasks},label={output:pse},escapeinside={(*@}{@*)}]
		sigma_ibo@sigma-ibo:~$ cat /sys/devices/system/cpu/isolated
		4
		sigma_ibo@sigma-ibo:~$ ps axHo psr,pid,lwp,args,policy,nice,rtprio | awk '$1 == 4'
		4      38      38 [cpuhp/4]                   TS    0      -
		4      39      39 [idle_inject/4]             FF    -     50
		4      40      40 [migration/4]               FF    -     99
		4      41      41 [ksoftirqd/4]               TS    0      -
		4      42      42 [kworker/4:0-events]        TS    0      -
		4      43      43 [kworker/4:0H-events_highpr TS  -20      -
		4     505     505 [irq/189-iwlwifi:queue_1]   FF    -     50
		4     519     519 [irq/203-iwlwifi:exception] FF    -     50
		4    3008    3008 [kworker/4:1H]              TS  -20      -
		4    3022    3022 [kworker/4:2-rcu_gp]        TS    0      -	  
\end{lstlisting}
\end{minipage}
\subsection{Interrupt Requests Handling}\label{sec:irq_handling}
Once the CPUs were isolated, Interrupt Requests Handling was the next step. Interrupt Requests are used to send a signal to the CPU, prompting it to 'interrupt' its current task and divert its attention to another task. This allows hardware devices to communicate with the CPU through frequent context switches, which can lead to performance degradation, especially in high-performance computing or real-time scenarios. To mitigate this, the IRQs needed to be removed from the isolated CPUs. This was done by manipulating a file in the proc filesystem, namely \texttt{/proc/irq/<IRQ>/smp\_affinity}. The value in the \texttt{smp\_affinity} file is a bitmask in hexadecimal format. Each bit in this mask corresponds to a CPU in the system. The least significant bit (LSB) on the right corresponds to the first CPU (CPU0), and the significance increases towards the left until CPU19. In a system with 20 CPUs, if every CPU was reserved for one IRQ, the value for \texttt{smp\_affinity} would be FFFFF. The script in code \ref{script:smp_affinity} was written to check and log the distribution of Interrupt Requests across each CPU in Salamander 4.

@@

\noindent By changing the values of the \texttt{smp\_affinity} files of the respective IRQs, the assignment of IRQs was controlled so that they would not be handled by the isolated CPU. This reduced the interruptions caused by IRQs and the isolated CPUs were able to focus more on their assigned tasks.
\subsection{Disable dynamic frequency scaling}
\subsection{Disable RT throttling}
\subsection{No unexpected RT processes are running on your system}
\subsection{IRQ affinity}
\subsection{RCU CPU offloading}
\subsection{Suppress rcu cpu stall}
\subsection{Maybe?}
\subsection{Start QEMU normally and give all QEMU threads rt-priority} 
\subsection{Kill all running user processes}
\subsection{Set CPU Affinity for systemd services}
\subsection{Cache Isolation for CPU and GPU}
\subsection{Set CPU Affinity of IRQ thread to CPU 0}
\subsection{Set Device Driver Work Queue to CPU 0}
\subsection{Disable Machine Check}
\subsection{Stop Certain Services}
\section{QEMU-KVM Configurations}
\subsection{Tune lapic timer advance}
\subsection{Set QEMU options for real-time VM}
\subsection{Set CPU affinity and scheduling policy of QEMU CPU threads}
\subsection{Passthrough PCI devices into the VM}

\clearpage
\section{Guest OS Configurations}
\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualisation},label={script:qemu_configuration_optimizations}]
    #!/bin/sh

@@

\chapter{To Include}\label{cha:to_include}
\section{Latency Comparison}
In the initial phase, a comparative latency analysis was conducted between the hardware version and the virtualized version of Salamander 4. For this purpose, the latency tool of the Xenomai test suite was used. The latency was measured under two conditions, idle and CPU-stressed. The goal was to optimize the latency of the virtualisation of Salamander 4 OS to closely match that of the bare metal version.
